Partial Identifiability

### 2016-07-26 identify/adjust/adjust.tex
makes data generating processes explicit
more on definitions of causal effect

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{tikz} 
\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[#1\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\N}[2]{\mathcal{N}\left({#1}\;,\;{#2}\right)}

@@

\begin{figure}[H]
	\centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 %nodes
\node[main node](2){$X$};
\node[main node, above right=of 2](3){$Z$};
\node[main node, below right=of 3](4){$Y$};
\node[latent, above=of 2](5){$\varepsilon_x$};
\node[latent, above=of 4](6){$\varepsilon_y$};
 \path[every node/.style={font=\tiny}]
    (3) edge (2) edge (4)
    (2) edge (4)
    (5) edge (2)
    (6) edge (4);
\end{tikzpicture}
\end{figure}

@@

\eq {
P(Y|do(X=x)) = \sum_{z \in \mathcal{Z}}P(Y|X=x,Z=z)P(Z)
}

@@

\eq{
ATE = E(Y|do(X=1)) - E(Y|do(X=0))
}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{dsfont}

\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{tikz} 
\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[2]{\mathbb E_{#1}\left[#2\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\N}[2]{\mathcal{N}\left({#1}\;,\;{#2}\right)}

@@

\begin{figure}[H]
    \begin{subfigure}[b]{0.45\textwidth}
	\centering    
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 %nodes
\node[main node](2){$X$};
\node[main node, above right=of 2](3){$Z$};
\node[main node, below right=of 3](4){$Y$};
\node[latent, above=of 2](5){$\varepsilon_x$};
\node[latent, above=of 4](6){$\varepsilon_y$};
 \path[every node/.style={font=\tiny}]
    (3) edge (2) edge (4)
    (2) edge (4)
    (5) edge (2)
    (6) edge (4);
\end{tikzpicture}
        \caption{prior to intervention}
        \label{fig:before}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
    \centering
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 %nodes
\node[main node](2){$x$};
\node[main node, above right=of 2](3){$Z$};
\node[main node, below right=of 3](4){$Y$};
\node[latent, above=of 4](6){$\varepsilon_y$};
 \path[every node/.style={font=\tiny}]
    (3) edge (4)
    (2) edge (4)
    (6) edge (4);
\end{tikzpicture}
        \caption{after $do(X=x)$}
        \label{fig:after}
    \end{subfigure}

    \caption{Data generating process}\label{fig:causalmodels}
    \label{fig:adjustment}
\end{figure}

The idea of causal inference is to use data generated by one process (prior to an intervention) to infer characteristics of another process (the process after intervention). Causal graphical model how an intervention changes the process. For each node with incoming links there is a (deterministic) function that determines its value. We also implicitly assume that these functions and the distributions of all variables without incoming links does not change as a consequence of the intervention. The specific case of causal estimation by adjusting for known confounders is shown in figure \ref{fig:adjustment}.

@@

\eq {
P(Y|do(X=x))= P(Y) \text{, in graph \ref{fig:after}} = \sum_{z \in \mathcal{Z}}P(Y|X=x,Z=z)P(Z) \text{, in graph \ref{fig:before}}
}

@@

\eq{
ITE &= Y|do(X=1),Z \leftarrow \text{ individual treatment effect, defined only if there's no unobserved noise $\epsilon_y$}\\
PTE(Z) &= \E{\epsilon_y}{Y|do(X=1),Z}-\E{\epsilon_y}{Y|do(X=0),Z} \leftarrow \text{ personalized treatment effect}\\
ATE &= \E{Z,\epsilon_y}{Y|do(X=1)} - \E{Z,\epsilon_y}{Y|do(X=0)} = \E{Z}{ITE(Z)} \leftarrow \text{ average treatment effect}
}

@@

\subsubsection*{Counterfactuals}
Let $Y = f_y(X,Z,\epsilon_y)$. Suppose we have a sample $\set{x_i = 1,z_i,y_i}$ and corresponding unobserved $\set{\epsilon_{x,i},\epsilon_{y,i}}$ from the model in figure \ref{fig:before}. The counterfactual $Y_1$, the value of y that would have been observed had $x=1$, is $y_i$ (since $x$ did equal 1). $Y_0$ is the value of y that would have been observed had $x=0$. Even if we know $f_y$, we cannot in general determine $Y_0$ because we did not observe $\epsilon_y$.

So now we have one thing that we know with certainty and another about which we could say something about in expectation. Does defining treatment effects in terms of counterfactuals in this way have lower noise - than in the other way because we are taking expectations only with respect to one of the parts?
